TMR: Detecting Structure in Graphical Data

### 2021-01-18 interim-report/interim-report.tex
More work

%opening
\title{Detecting Structure in Graphical Data}
\author{Lawrence Tray \\ Ioannis Kontoyiannis}

%packages
\usepackage[margin=0.5in]{geometry}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{hyperref}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{mathtools}
\usepackage{parskip}
\usepackage{fancyvrb}
\usepackage{bbm}

@@

%custom commands
\newcommand{\dft}{\mathcal{F}}
\newcommand{\idft}{\mathcal{F}^{-1}}
\newcommand{\Xcal}{\mathcal{X}}
\newcommand{\Dcal}{\mathcal{D}}
\newcommand{\cmplx}{\mathbb{C}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\Gaussian}{\mathcal{N}}
\newcommand{\Gcal}{\mathcal{G}}
\newcommand{\Vcal}{\mathcal{V}}
\newcommand{\Ecal}{\mathcal{E}}
\newcommand{\pbold}{\boldsymbol{p}}
\newcommand{\sigmabold}{\boldsymbol{\sigma}}
\newcommand{\lik}{\mathcal{L}}
\newcommand{\kl}{\mathcal{D}}
\newcommand{\Integers}{\mathbb{Z}}
\newcommand{\SBM}{\textrm{SBM}}
\newcommand{\iid}{\stackrel{iid}{\sim}}
\newcommand{\one}{\mathbbm{1}}
\newcommand{\figwidth}{0.55\linewidth}

@@

\begin{abstract}
So the agenda for today. I will take you through my exploratory work around the field in rough chronological order. We’ll start off with the fundamentals of hypothesis testing. I will then introduce you to the most commonly used graphical model and talk about the ways of verifying structure in a graph with labelled nodes. However, that is not sufficient as we may want to detect structure in an unlabelled graph. Finally I’ll talk about the future direction of the project.
\end{abstract}

@@

\section{Verifying Structure}

@@

Furthermore, we spot the following property $f(w, e_1, m_1) \cdot f(w, e_2, m_2) = f(w, e_1 + e_2, m_1 + m_2)$ or in other words, the function $f$ is linear in its second and third arguments given the same first argument. As such we can manipulate equation \ref{eqn:test-statistic-start} greatly to give:
%
\begin{align}
	t_n &= \log \frac
	{
		\max_{W_{\alpha} \neq W_{\beta}}(f (W_\alpha, E_\alpha, M_\alpha) \cdot f (W_\beta, E_\beta, M_\beta))
	}
	{
		\max_{W_\alpha = W_\beta} (f (W_\alpha, E_\alpha, M_\alpha) \cdot f (W_\beta, E_\beta, M_\beta))
	} \nonumber \\
	&= \log \frac{
		\max_p f(p, E_\alpha, M_\alpha) \cdot \max_q f(q, E_\alpha, M_\alpha)
	}{
		\max_r f(r, E_\alpha + E_\beta, M_\alpha + M_\beta)
	} \nonumber \\
	&= \log \frac{
		f(\hat{p}, E_\alpha, M_\alpha) \cdot f(\hat{q}, E_\alpha, M_\alpha)
	}{
		f(\hat{r}, E_\alpha + E_\beta, M_\alpha + M_\beta)
	} \nonumber \\
	&= \log \frac{f(\hat{p}, E_\alpha, M_\alpha)}{f(\hat{r}, E_\alpha, M_\alpha)} + \log \frac{f(\hat{q}, E_\beta, M_\beta)}{f(\hat{r}, E_\beta, M_\beta)}
\end{align}

@@

\begin{lemma}
	With $f$ defined as in equation \ref{eqn:f-defn}, $0 \leq e \leq m$ and $r \in [0, 1]$ it holds that $\log \frac{f(e/m, e, m)}{f(r, e, m)} = m \cdot \kl \left( \textrm{Bern}(e/m) || \textrm{Bern}(r) \right)$ where $\kl(g || h)$ is the Kullback-Leibler divergence between two probability mass functions $g, h: \Xcal \mapsto [0, 1]$ defined in discrete space as $\kl(g || h) \coloneqq \sum_{x \in \Xcal} g(x) \log \frac{g(x)}{h(x)}$ and $\textrm{Bern(p)}$ denotes the Bernoulli p.m.f with parameter $p$.
	\label{lem:kl-div}
\end{lemma}

Proving lemma \ref{lem:kl-div} is simply a case of algebraic manipulation:
%
\begin{align}
	\log \frac{f(e/m, e, m)}{f(r, e, m)} &= e \cdot \log \frac{e/m}{r} + (m-e) \cdot \log \frac{1- e/m}{1 - r} \nonumber \\
	&= m \cdot \left( (e/m) \cdot \log \frac{e/m}{r} + (1 - e/m) \cdot \log \frac{1- e/m}{1 - r} \right) \nonumber \\
	&= m \sum_{x \in \{0, 1\}} \textrm{Bern}(x; e/m) \cdot \log \frac{\textrm{Bern}(x; e/m)}{\textrm{Bern}(x; r)} \nonumber \\
	&= m \kl \left( \textrm{Bern}(e/m) || \textrm{Bern}(r) \right) \quad \therefore \textrm{QED}
\end{align}

@@

\begin{lemma}
	For $p \approx r$ then 
\end{lemma}

\begin{theorem}
	We posit that $(X, \Gcal) \sim \textrm{SBM}(n, p, W)$. Given the realised graph and class labels $(X, \Gcal)$ we can perform a hypothesis test on parameters $W_\alpha$ and $W_\beta$ of the connectivity matrix $W$.
	%
	\begin{align*}
	H_0:& \quad W_{\alpha} = W_{\beta} \\
	H_1:& \quad W_{\alpha} \neq W_{\beta}
	\end{align*}
	If the log-likelihood ratio test statistic $t_n$ is computed as in equation \ref{eqn:test-stat-compute-form}, repeated here:
	%
	\begin{equation*}
		t_n \coloneqq  M_\alpha \cdot \kl\left( \textrm{Bern}(\hat{p}) || \textrm{Bern}(\hat{r})\right) + 
		M_\beta \cdot \kl\left( \textrm{Bern}(\hat{q}) || \textrm{Bern}(\hat{r})\right)
	\end{equation*}

@@

\begin{corollary}
	We can also use a slightly simpler test-statistic $z_n$ albeit with some loss of generality. If we define $z_n$ to be:
	%
	\begin{equation*}
		z_n \coloneqq \sqrt{\frac{M_\alpha M_\beta}{\hat{r}(1-\hat{r})(M_\alpha + M_\beta)}} \left( \hat{p} - \hat{q} \right)
	\end{equation*}

	With the symbols retaining their previous meanings. Then we show that under the null $H_0$, as $n \rightarrow \infty$ then $z_n \sim \Gaussian(0, 1)$. Meaning that, we can construct a similar test to reject $H_0$ at the $100(\zeta)\%$ confidence level if and only if $|z_n| \geq \phi^{-1}(\zeta)$, where $\phi^{-1}$ is the standard inverse Gaussian cdf on magnitude satisfying $Pr(|Y| \leq \phi^{-1}(\zeta)) = \zeta$ given $Y \sim \Gaussian(0, 1)$
\end{corollary}

\section{Early results}
\subsection{Social Networks}

@@

\section{Detecting Structure}

@@

\begin{definition}
	For $(X, \Gcal) \sim \textrm{SBM}(m, p, W)$ the following recovery requirements are solved if there exists an algorithm which takes $\Gcal$ and input and outputs a vector of classifications $\hat{X} = \hat{X}(\Gcal)$ such that as $n \rightarrow \infty$:
	\begin{itemize}
		\item \textbf{Exact recovery:} $Pr\{A(X, \hat{X}) = 1\} = 1 - o(1)$ correct classification recovered almost surely
		\item \textbf{Almost exact recovery:} $Pr\{A(X, \hat{X}) = 1 - o(1)\} = 1 - o(1)$ vanishing fraction misclassified almost surely
		\item \textbf{Partial recovery:} $Pr\{\tilde{A}(X, \hat{X}) \geq \alpha\} = 1 - o(1), 1 > \alpha > 1/k$ better than choosing from uniform prior
		\item \textbf{Weak recovery:} $Pr\{\tilde{A}(X, \hat{X}) \geq 1/k + \epsilon \} = 1 - o(1), \epsilon > 0$ marginally better than choosing from uniform prior
		\item \textbf{No recovery:} graph provides no information as to class labels
	\end{itemize}
	Where $o(1)$ denotes a family of functions that tend to 0 as $n \rightarrow \infty$. The regime $\Gcal$ occupies depends on the parameters of the SBM. Each recovery regime is weaker than the one above it so Exact recovery implies weak recovery is possible..
\end{definition}

We choose to focus on weak recovery as it is the least strict of all regimes and in algorithms for weak recovery can solve stricter recovery regimes too. One of the most promising weak recovery algorithms is called Acyclic Belief Propagataion (ABP). We skim over the details here and merely represent our form of the algorithm, adapted from \cite{Linear-ABP}. ABP(r, T) on a graph $\Gcal$ with vertex set $\Vcal = \Vcal(\Gcal)$ and edge set $\Ecal = \Ecal(\Gcal)$:

\begin{enumerate}
	\item Initialise messages for $(v, v')$:
	\subitem $y^{(0)}_{v' \rightarrow v} \leftarrow \Gaussian(0, 1)$
	\item Iterate for $1 \leq t \leq T$ and for $(v, v') \in \Ecal$:
	\subitem compute average $s^{(t-1)} \leftarrow \frac{1}{2|E|} \sum_{(v, v') \in \Ecal} y^{(t-1)}_{v' \rightarrow v}$
	\subitem recentre messages $z^{(t-1)}_{v' \rightarrow v} \leftarrow y^{(t-1)}_{v' \rightarrow v} - s^{(t-1)}$ 
	\subitem sum incoming $y^{(t)}_{v' \rightarrow v} \leftarrow \sum_{(v', v'') \in \Ecal \setminus \{v\}}z^{(t-1)}_{v'' \rightarrow v'}$
	\subitem if $(v''' \rightarrow v \rightarrow v')$ on cycle of length $r' \leq r$ then correct:
	\subsubitem $y^{(t)}_{v' \rightarrow v} \leftarrow y^{(t)}_{v' \rightarrow v} - \sum_{(v, v'') \in \Ecal \setminus \{v', v'''\}}z^{(t-r')}_{v'' \rightarrow v'}$
	\item Assignment, for all $v \in V$:
	\subitem Sum incoming $y_v^{(T)} = \sum_{(v, v') \in \Ecal} y^{(t)}_{v' \rightarrow v}$
	\subitem Assign labels $\sigma_v = 1$ if $y_v^{(T)} > 0$ and $0$ otherwise
\end{enumerate}

We implement this algorithm in Python and apply it to the Facebook Egonets described earlier.

\section{Verifying Structure}
\subsection{Single Sample against known mean}
We start with the simple case of determining whether or not a coin is fair. Each coin flip can be represented by a random variable with a Bernoulli distribution $X_i \sim Bern(p)$. Each coin can result in either a Tails or a Heads denoted by $X_i \in \{0, 1\}$. We toss the coin $n$ times leading us to a set $\left\{ X_i \right\}_{i=1}^{n}$. We wish to test the null hypothesis $p=p_0$ against the alternative. We shall keep the $p_0$ notation for generality, though for a fair coin we require $p_0=1/2$.
%
\begin{align*}
H_0:& \quad p = p_0 \\
H_1:& \quad p \neq p_0
\end{align*}
%
We denote the number of heads with the random variable $K \coloneqq \sum_{i=1}^{n} X_i \sim Bern(p, n)$. For a particular experiment we observe $K=k$. We employ a standard likelihood ratio test. The test statistic $t_n$ is calculated as the log-likelihood ratio of observing $K=k$ under $H_1$ and $H_0$.
%
\begin{equation}
t_n \coloneqq \log \frac{\lik(H_1)}{\lik(H_0)}
= \log \frac{\max_p P(K=k|p)}{P(K=k|p=p_0)}
\end{equation} 
%
We note that since $K$ is distributed as a binomial, $P(K=k|p)=\binom{n}{k}p^k(1-p)^{n-k}$. If we can vary $p$, this probability is maximised for $p=\hat{p} \coloneqq k/n$. Therefore, the test statistic is given by.
%
\begin{equation}
t_n = \log \frac{\binom{n}{k} \hat{p}^k (1 - \hat{p})^{n-k}}{\binom{n}{k} p_0^k (1 - p_0)^{n-k}}
=  \log \frac{ \hat{p}^k (1 - \hat{p})^{n-k}}{ p_0^k (1 - p_0)^{n-k}}
\end{equation}
%
The combinatoric term $\binom{n}{k}$ cancels out. This implies that the order in which the heads land does not matter when determining the fairness of the coin. We can work the above expression into a more usable form:
%
\begin{align}
t_n =& k \log \frac{\hat{p}}{p_0} + (n-k) \log \frac{1 - \hat{p}}{1 - p_0} \nonumber \\
=& n \left( \hat{p} \log \frac{\hat{p}}{p_0} + (1-\hat{p}) \log \frac{1 - \hat{p}}{1-p_0} \right) \nonumber \\
=& n \kl \left(Bern(\hat{p}) || Bern(p_0) \right)
\end{align}
%
Where $\kl(f | g) \coloneqq \sum_{x \in \Xcal} f(x) \log \frac{f(x)}{g(x)}$ is the Kullback-Leibler divergence between two arbitrary probability mass functions $f$ and $g$. This is also called the relative entropy. The KL divergence has the property that:
%
\begin{equation}
\kl (f || g) \geq 0 \quad \text{with equality iff} \quad f(x) = g(x) \quad \forall x \in \Xcal
\end{equation}
%
We can exploit this result for the case that $f \approx g$ to obtain a simplified expression for the KL divergence. We begin by defining $\delta (x) \coloneqq f(x) - g(x)$. We are interested in the region where $\delta$ is small. We start by substituting for $f=\delta + g$ and then taking the Taylor expansion of $\log 1+x$.
%
\begin{align*}
\kl(f||g) &= \sum_{x \in \Xcal} (\delta + g) \log \left(1 + \frac{\delta}{g} \right) \\
&= \sum_{x \in \Xcal} (\delta + g) \left( \frac{\delta}{g} - \frac{\delta^2}{2g^2} + O(\delta^3) \right) \\
&= \sum_{x \in \Xcal} \delta + \frac{1}{2} \sum_{x \in \Xcal} \frac{\delta^2}{g} + O(\delta^3) \\
&= \frac{1}{2} \sum_{x \in \Xcal} \frac{\delta^2}{g} + O(\delta^3) \\
&= \frac{1}{2} \chi^2(f||g) + O(\delta^3)
\end{align*}
%
Where the summation over $\delta$ evaluates to 0 because $\delta$ is the difference of two valid p.m.f's which each sum to 1 over $x \in \Xcal$. We are able to neglect the $O(\delta^3)$ terms for $f$ very close to $g$ we shall see what this means later. $\chi^2(f||g)$ is known as the chi-squared distance between two distributions and is defined simply as $\chi^2(f||g) \coloneqq \sum_{x \in \Xcal} (f-g)^2/g$. We now investigate the chi-squared divergence for $f = Bern(p)$ and $g = Bern(q)$.
%
\begin{align*}
\chi^2(Bern(p)||Bern(q)) &= \frac{(p-q)^2}{q} + \frac{((1-p)-(1-q))^2}{1-q} \\
&= \frac{(p-q)^2}{q(1-q)} \\
&= \left(\frac{p-q}{\sqrt{q(1-q)}}\right)^2
\end{align*}
%
Now we can exploit these results to get a workable expression for the test statistic $t_n$.
%
\begin{align*}
t_n &= n \kl \left(Bern(\hat{p}) || Bern(p_0)\right) \\
	&= \frac{n}{2} \chi^2 \left(Bern(\hat{p}) || Bern(p_0) \right) + nO(\delta^3) \\
	&= \frac{1}{2} \left( \frac{\hat{p} - p_0}{\sqrt{p_0(1-p_0)/n}} \right)^2 + \epsilon
\end{align*}
%

>>>>

%opening
\title{TMR: Detecting Structure in Graphical Data}
\author{Lawrence Tray \\ Ioannis Kontoyiannis}

%packages
\usepackage[margin=0.5in]{geometry}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{hyperref}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{mathtools}
\usepackage{parskip}
\usepackage{fancyvrb}
\usepackage{bbm}
\usepackage{algorithm}% http://ctan.org/pkg/algorithm
\usepackage{algpseudocode}% http://ctan.org/pkg/algorithmicx

@@

%custom commands
\newcommand{\dft}{\mathcal{F}}
\newcommand{\idft}{\mathcal{F}^{-1}}
\newcommand{\Xcal}{\mathcal{X}}
\newcommand{\Dcal}{\mathcal{D}}
\newcommand{\cmplx}{\mathbb{C}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\Gaussian}{\mathcal{N}}
\newcommand{\Gcal}{\mathcal{G}}
\newcommand{\Vcal}{\mathcal{V}}
\newcommand{\Ecal}{\mathcal{E}}
\newcommand{\pbold}{\boldsymbol{p}}
\newcommand{\sigmabold}{\boldsymbol{\sigma}}
\newcommand{\lik}{\mathcal{L}}
\newcommand{\kl}{\mathcal{D}}
\newcommand{\Integers}{\mathbb{Z}}
\newcommand{\SBM}{\textrm{SBM}}
\newcommand{\iid}{\stackrel{iid}{\sim}}
\newcommand{\one}{\mathbbm{1}}
\newcommand{\figwidth}{0.4\linewidth}

@@

\begin{abstract}
This is the Technical Milestone Report for the II-B project: detecting structure in graphical data. We produce vast quantities of graphical data every day and yet the techniques used to analyse their structure are still in their infancy. We start by defining the Stochastic Block Model (SBM) widely used in academia and develop a theorem to verify structure in a given labelled graph.
\end{abstract}

@@

\section{Verifying Structure}
\subsection{Theory}

@@

Furthermore, we spot the following property $f(w, e_1, m_1) \cdot f(w, e_2, m_2) = f(w, e_1 + e_2, m_1 + m_2)$ or in other words, the function $f$ is linear in its second and third arguments given the same first argument. As such we can manipulate equation \ref{eqn:test-statistic-start} greatly to give:
%
\begin{align}
	t_n &= \log \frac
	{
		\max_{W_{\alpha} \neq W_{\beta}}(f (W_\alpha, E_\alpha, M_\alpha) \cdot f (W_\beta, E_\beta, M_\beta))
	}
	{
		\max_{W_\alpha = W_\beta} (f (W_\alpha, E_\alpha, M_\alpha) \cdot f (W_\beta, E_\beta, M_\beta))
	} \nonumber \\
	&= \log \frac{
		\max_p f(p, E_\alpha, M_\alpha) \cdot \max_q f(q, E_\alpha, M_\alpha)
	}{
		\max_r f(r, E_\alpha + E_\beta, M_\alpha + M_\beta)
	} \nonumber \\
	&= \log \frac{f(\hat{p}, E_\alpha, M_\alpha)}{f(\hat{r}, E_\alpha, M_\alpha)} + \log \frac{f(\hat{q}, E_\beta, M_\beta)}{f(\hat{r}, E_\beta, M_\beta)}
\end{align}

@@

\begin{lemma}[KL divergence]
	With $f$ defined as in equation \ref{eqn:f-defn}, $0 \leq e \leq m$ and $r \in [0, 1]$ it holds that:
	%
	\begin{equation*}
		\log \frac{f(e/m, e, m)}{f(r, e, m)} = m \cdot \kl \left( \textrm{Bern}(e/m) || \textrm{Bern}(r) \right)
	\end{equation*}

	where $\kl(g || h)$ is the Kullback-Leibler divergence between two probability mass functions $g, h: \Xcal \mapsto [0, 1]$. $\kl$ is defined in discrete space as $\kl(g || h) \coloneqq \sum_{x \in \Xcal} g(x) \log \frac{g(x)}{h(x)}$ and $\textrm{Bern(p)}$ denotes the Bernoulli p.m.f with parameter $p$.
	\label{lem:kl-div}
\end{lemma}

Proving lemma \ref{lem:kl-div} is simply a case of algebraic manipulation:
%
\begin{align}
	\log \frac{f(e/m, e, m)}{f(r, e, m)} &= e \cdot \log \frac{e/m}{r} + (m-e) \cdot \log \frac{1- e/m}{1 - r} \nonumber \\
	&= m \cdot \left( (e/m) \cdot \log \frac{e/m}{r} + (1 - e/m) \cdot \log \frac{1- e/m}{1 - r} \right) \nonumber \\
	&= m \kl \left( \textrm{Bern}(e/m) || \textrm{Bern}(r) \right) \quad \therefore \textrm{QED}
\end{align}

@@

However, we must simplify further still.
%
\begin{lemma}[$\chi^2$ approximation]
	\label{lem:kl-approx-chi}
	For $g(x) \approx h(x)$ then $\kl(g || h) \approx \frac{1}{2} \chi^2(g || h)$ where $\chi^2$ is the chi-squared distance between two distributions defined as $\chi^2(g || h) \coloneqq \sum_{x \in \Xcal} \frac{(g(x)-h(x))^2}{h(x)}$ 
\end{lemma}

The proof requires defining $\delta(x) \coloneqq g(x) - h(x)$ and taking a Taylor expansion of $\log 1 + x$:
%
\begin{align*}
	\kl(g||h) &= \sum_{x \in \Xcal} (\delta + h) \log \left(1 + \frac{\delta}{g} \right) 
	= \sum_{x \in \Xcal} (\delta + h) \left( \frac{\delta}{h} - \frac{\delta^2}{2h^2} + O(\delta^3) \right)  \\
	&= \sum_{x \in \Xcal} \delta + \frac{1}{2} \sum_{x \in \Xcal} \frac{\delta^2}{h} + O(\delta^3) 
	= \frac{1}{2} \sum_{x \in \Xcal} \frac{\delta^2}{h} + O(\delta^3) \\
	&= \frac{1}{2} \chi^2(g||h) + O(\delta^3) \approx \frac{1}{2} \chi^2(h||g) \quad \therefore \textrm{QED}
\end{align*}
%
Where the summation over $\delta$ evaluates to 0 because $\delta$ is the difference of two valid p.m.f's which each sum to 1 over $x \in \Xcal$. We are able to neglect the $O(\delta^3)$ terms for $g$ very close to $h$. The chi-squared distance between two Bernoullis for $g = Bern(p)$ and $h = Bern(q)$ is simply:
%
\begin{equation}
\chi^2(\textrm{Bern}(p) || \textrm{Bern}(q)) = \left( \frac{p-q}{\sqrt{q(1-q)}} \right)^2
\end{equation}

We can apply these results to equation \ref{eqn:test-stat-compute-form} to obtain an approximate expression for $t_n$ under the null for large $n$ such that $\hat{p}$, $\hat{q}$ and $\hat{r}$ are all close meaning Lemma \ref{lem:kl-approx-chi} is valid, giving:
%
\begin{align}
	t_n &\approx \frac{1}{2 \hat{r}(1 - \hat{r})} \left( 
	M_\alpha \left(\hat{p} - \hat{r}\right)^2 +
	M_\beta \left(\hat{q} - \hat{r} \right)^2 
	\right) \nonumber \\
	&= \frac{1}{2 \hat{r}(1-\hat{r})} \left( 
	M_\alpha \left(\frac{M_\alpha (\hat{p} - \hat{q})}{M_\alpha + M_\beta} \right)^2 +
	M_\beta \left(\frac{M_\beta (\hat{q} - \hat{p})}{M_\alpha + M_\beta} \right)^2
	\right) \nonumber \\
	&= \frac{1}{2} \left( \frac{\hat{p} - \hat{q}}{\sqrt{\hat{r}(1- \hat{r})(1/M_\alpha + 1/M_\beta)}} \right)^2
	= \frac{1}{2} z_n^2
\end{align}

Through application of the Central Limit Theorem (CLT) and Slutsky's Theorem it can be shown that as $n \rightarrow \infty$ under the null $H_0$ the term in brackets on the final line is distributed like a standard Gaussian $z_n \sim \Gaussian(0,1)$. This is because under the same assumptions, assuming that $W_\alpha = W_\beta = \mu$, by the CLT $\hat{p} \sim (\mu, \mu(1 - \mu)/ M_\alpha)$, $\hat{q} \sim (\mu, \mu(1 - \mu)/ M_\beta)$ and $\hat{p} \sim (\mu, \mu(1 - \mu)/ (M_\alpha + M_\beta))$. Therefore, the whole test-statistic is distributed as $t_n \sim \frac{1}{2} \chi^2_1$ by the definition of the $\chi^2_p$ distribution: $\chi^2_p \sim \sum_{i=1}{p} Z_i^2$ for $Z_i \sim \Gaussian(0,1)$. This leads to the following theorem:

\begin{theorem}[$\chi^2$ test]
	For $(X, \Gcal) \sim \textrm{SBM}(n, p, W)$, given the realised graph and class labels $(X, \Gcal)$ we can perform a hypothesis test on parameters $W_\alpha$ and $W_\beta$ of the connectivity matrix $W$.
	%
	\begin{align*}
	H_0:& \quad W_{\alpha} = W_{\beta} \\
	H_1:& \quad W_{\alpha} \neq W_{\beta}
	\end{align*}
	If the log-likelihood ratio test statistic $t_n$ is computed as in equation \ref{eqn:test-stat-compute-form}, repeated here:
	%
	\begin{equation*}
		t_n \coloneqq  M_\alpha \cdot \kl\left( \textrm{Bern}(\hat{p}) || \textrm{Bern}(\hat{r})\right) + 
		M_\beta \cdot \kl\left( \textrm{Bern}(\hat{q}) || \textrm{Bern}(\hat{r})\right)
	\end{equation*}

@@

\begin{corollary}[z-test]
	We can also use a slightly simpler test-statistic $z_n$ albeit with some loss of generality. If we define $z_n$ to be:
	%
	\begin{equation*}
		z_n \coloneqq \frac{\hat{p} - \hat{q}}{\sqrt{\hat{r}(1- \hat{r})(1/M_\alpha + 1/M_\beta)}}
	\end{equation*}

	With the symbols retaining their previous meanings. Under the null $H_0$, as $n \rightarrow \infty$ then $z_n \sim \Gaussian(0, 1)$. Meaning that, we can construct a similar test to reject $H_0$ at the $100(\zeta)\%$ confidence level if and only if $|z_n| \geq \phi^{-1}(\zeta)$, where $\phi^{-1}$ is the standard inverse Gaussian cdf on magnitude satisfying $Pr(|Y| \leq \phi^{-1}(\zeta)) = \zeta$ given $Y \sim \Gaussian(0, 1)$
\end{corollary}

\subsection{Early results}

@@

\section{Detecting Structure}
\subsection{Weak recovery}

@@

\begin{definition}[Recovery regimes]
	for $(X, \Gcal) \sim \textrm{SBM}(m, p, W)$ the following recovery requirements are solved if there exists an algorithm which takes $\Gcal$ and input and outputs a vector of classifications $\hat{X} = \hat{X}(\Gcal)$ such that as $n \rightarrow \infty$:
	\begin{itemize}
		\item \textbf{Exact recovery:} $Pr\{A(X, \hat{X}) = 1\} = 1 - o(1)$ correct classification recovered almost surely
		\item \textbf{Almost exact recovery:} $Pr\{A(X, \hat{X}) = 1 - o(1)\} = 1 - o(1)$ vanishing fraction misclassified almost surely
		\item \textbf{Partial recovery:} $Pr\{\tilde{A}(X, \hat{X}) \geq \alpha\} = 1 - o(1), 1 > \alpha > 1/k$ better than choosing from uniform prior
		\item \textbf{Weak recovery:} $Pr\{\tilde{A}(X, \hat{X}) \geq 1/k + \epsilon \} = 1 - o(1), \epsilon > 0$ marginally better than choosing from uniform prior
		\item \textbf{No recovery:} graph provides no information as to class labels
	\end{itemize}
	Where $o(1)$ denotes a family of functions that tend to 0 as $n \rightarrow \infty$. The regime $\Gcal$ occupies depends on the parameters of the SBM. Each recovery regime is weaker than the one above it so Exact recovery implies weak recovery is possible..
\end{definition}

We choose to focus on weak recovery as it is the least strict of all regimes. An alternative statement of weak recovery is given below:

\begin{definition}[Weak recovery]
	\label{defn:weak-recovery}
	for  $(X, \Gcal) \sim \textrm{SBM}(n, p, W)$ weak recovery is solved if there exists $\epsilon > 0$ and a certain choice of indices $i, j \in \{1, 2 \dots k\}$ and an algorithm that takes as input $\Gcal$ and outputs a partition of the vertex set $\Vcal$ into two distinct sets $S$ and $S^C$ (detected communities 1 and 2 respectively) such that:
	\begin{equation*}
		Pr\left\{ \frac{|\Omega_i \cap S|}{|\Omega_i|} - \frac{|\Omega_j \cap S|}{|\Omega_j|} \geq \epsilon \right\} = 1 - o(1)
	\end{equation*}

	Bearing in mind the definition of $\Omega_i = \Omega_i(X) \coloneqq \{v \in \Vcal : X_v = i\}$. In other words, we require the partition $S \subset \Vcal$ to cut across communities such that the fraction of a particular community in partition $S$ with respect to the whole community is different for a choice of two communities.  
\end{definition}

One of the most promising weak recovery algorithms is called Acyclic Belief Propagataion (ABP). This is a message-passing algorithm that solves weak recovery in the sense of definition \ref{defn:weak-recovery}. Algorithm \ref{alg:lin-abp} is heavily adapated from Abbe and Sandon \cite{Linear-ABP}. The original paper goes into much more detail but it works on the premise that one can infer the community of a particular vertex by taking a majority vote among the  of all its neighbours. The algorithm is non-determinsitic and requires the vertex set $\Vcal = \Vcal(\Gcal)$ and edge set $\Ecal = \Ecal(\Gcal)$ as well as two hyper-parameters $r$ (the maximum cyclic length to correct for) and $T$ (the number of iterations):

%
\begin{algorithm}
	\caption{Linearised Acyclic Belief Propagation (ABP)}
	\label{alg:lin-abp}
	\begin{algorithmic}
		\Function{ABP}{$\Ecal$, $\Vcal$, r, T}
		\State $y^{(0)}_{v' \rightarrow v} \gets \Gaussian(0, 1)$ \Comment{Initialise messages randomly}
		\For{$t \in \{1, 2 \cdots T\}$}
		\For{$(v, v') \in \Ecal$}
		\State $s^{(t-1)} \gets \frac{1}{2|E|} \sum_{(v, v') \in \Ecal} y^{(t-1)}_{v' \rightarrow v}$ \Comment{Compute average}
		\State $z^{(t-1)}_{v' \rightarrow v} \gets y^{(t-1)}_{v' \rightarrow v} - s^{(t-1)}$ \Comment{Recentre messages}
		\State $y^{(t)}_{v' \rightarrow v} \gets \sum_{(v', v'') \in \Ecal \setminus \{v\}}z^{(t-1)}_{v'' \rightarrow v'}$ \Comment{Sum incoming messages} \\
		\If {$\exists v''' \in \Vcal$ such that $(v''' \rightarrow v \rightarrow v')$ is on a cycle of length $r' \leq r$}
		\State $y^{(t)}_{v' \rightarrow v} \leftarrow y^{(t)}_{v' \rightarrow v} - \sum_{(v, v'') \in \Ecal \setminus \{v', v'''\}}z^{(t-r')}_{v'' \rightarrow v'}$ \Comment{Cyclic correction}
		\EndIf
		\EndFor
		\EndFor
		\State $\sigma_v = \sum_{(v, v') \in \Ecal} y^{(T)}_{v' \rightarrow v}$ \Comment{Sum incoming messages}
		\State $\hat{X}_v \gets 1$ if $\sigma_v > 0$ else $2$ \Comment{Hard assignment}
		\State \textbf{return} $\hat{X}_v$
		\EndFunction
	\end{algorithmic}
\end{algorithm}

We implement this algorithm in Python and apply it to the Facebook Egonets described earlier. We found that the hyperparameter settings $r=3$ and $T=5$ yielded good results. We define the normalised ABP output below:
%
\begin{equation}
	\tilde{\sigma}_v \coloneqq \begin{cases}
		\frac{\sigma_v}{|\sigma_{max}|} \textrm{ for } \sigma_j > 0 \\
		\frac{\sigma_v}{|\sigma_{min}|} \textrm{ for } \sigma_j \leq 0
	\end{cases}
\end{equation}

Where $\sigma_{max}$ is the largest positive value of $\sigma_v$ and $\sigma_{min}$ is the most negative. The normalised quantity $\tilde{\sigma}_v \in [-1, +1]$ with $0$ still being the midpoint that determines which detected community vertex $v$ is assigned to. We colour each vertex according to this quantity $\tilde{\sigma}_v$ and plot the graph on figure \ref{fig:abp-output-norm}. We see intuitively that ABP has worked. The blue cluster in the lower right is clearly part of one community and a great many vertices do not seem to belong to any cluster at all so their classification is uncertain (the colour is close to green).
%
\begin{figure}[!h]
	\centering
	\begin{subfigure}{0.4\linewidth}
		\includegraphics[width=\linewidth]{ego-0-abp.png}
		\caption{Normalised ABP output $\tilde{\sigma}_v$ colouring vertices}
		\label{fig:abp-output-norm}
	\end{subfigure}
	\begin{subfigure}{0.4\linewidth}
		\includegraphics[width=\linewidth]{ego-0-abp-by-gender.png}
		\caption{Vertices grouped by hard ABP assignments and coloured by gender. In assigned community 1 ($S$ - left) we have \{58, 122\} gender-77 and gender-78 respectively and in detected community 2 ($S^C$ - right) we have \{72, 96\}}
		\label{fig:abp-output-by-gender}
	\end{subfigure}
	\caption{Egonet-0 weak recovery results}
\end{figure}

We can evaluate formally whether weak recovery has succeeded by inspecting figure \ref{fig:abp-output-by-gender} and treating gender-77 and gender-78 as the two communities of interest:
%
\begin{equation}
	\frac{|\Omega_{77} \cap S|}{|\Omega_{77}|} = \frac{58}{58+72} = 0.446
	\quad \textrm{and} \quad
	 \frac{|\Omega_{78} \cap S|}{|\Omega_{78}|} = \frac{122}{122+96} = 0.560
\end{equation}

Clearly the fraction of each community in the partition $S$ differ substantially so we say that weak recovery is solved. The set $S$ contains a higher proportion of the overall 78-population than of the 77-population.

\subsection{From detection to inference}

We now have an algorithm (ABP) that partitions our graph into two sets that are most readily explained by an SBM with $k=2$ communities. We can now ask the question of which vertex labels most readily explain the separation. Therefore, we can determine which features have the largest impact on the graphical structure; this circumvents the binary resolution problem with the hypothesis testing approach.

This part of the project is still in its early stages so these are just preliminary results. For reference we plot the community fractions in the detected set $S$
\footnote{ABP is non-deterministic so each subplot is the result of a different run of ABP. Subfigures use the same partition $S$ but no exact comparison is possible between figures.}
on figure \ref{fig:abp-output-community-fractions}. It is interesting to note that if two bars on this plot differ substantially then we have satisfied the requirment for weak recovery as defined in \ref{defn:weak-recovery}. We necessarily remove the constraint on the community sets $\Omega_i$ to be disjoint as one vertex can have multiple features turned on. This can be modelled more precisely through a Mixed Membership Model but for now an SBM with $k=2$ communities suffices.
